---
title: "Bivariate Relationships Assignment"
output:
  html_document:
    highlight: tango
    theme: cerulean
css: ../css/customh5.css
---

# Step 1: Determine what variables for your two main constructs you want to graph based on your research topic.
You will need a mixture of categorical and quantitative variables for this assignment. 

* One association with a categorical explanatory and categorical response variable. (C ~ C)
* One association with a categorical explanatory and quantitative response variable. (Q ~ C)
* One association with a quantitative explanatory and binary categorical response variable. (B ~ Q)
* One association with a quantitative explanatory and quantitative response variable. (Q ~ Q)

If you do not have 1 or more of the combinations above you will do one of the specific types mentioned more than once.

_Recall the following definitions:_

* Explanatory variable: independent variable, covariate, goes along the x axis
* Response variable: dependent variable, outcome, goes along the y axis. 
* Model definition: y ~ x
    - C: categorical
    - B: binary categorical
    - Q: quantitative numeric / continuous
    
# Step 2: Recode variables as needed. 
If your response variable is categorical with many levels, you may want to collapse it down to fewer levels.  The example code below uses the `diamonds` data set from the `ggplot2` package. Here I combine levels `D` and `E` into a single category `DE`, `F` and `G` are combined into `FG`, and the remaining three `H`, `I`, and `J` are combined into a `HIJ` category. 

The code `car::recode()` means I am using the `recode()` function that is located in the `car` library. This is a shortcut to using a function inside a package without having to load that package. 


```{r}
library(ggplot2)
table(diamonds$color)

diamonds$col_cat <- car::recode(diamonds$color, "'D' = 'DE'; 'E' = 'DE'; 
                                     'F' = 'FG'; 'G' = 'FG';
                                     'H' = 'HIJ'; 'I' = 'HIJ'; 'J' = 'HIJ'")
table(diamonds$col_cat, diamonds$color)
```

If I simply need a binary indicator, the `factor()` function is more easily used. Recall that the first logical statement `diamonds$cut=="Ideal"` will result in a vector of `TRUE` which is treated as a 1, and `FALSE` which is treated as a 0. Then the `labels=` argument applies appropriate lables in ascending numerical order: 0 first, then 1. 

```{r}
table(diamonds$cut)
diamonds$ideal <- factor(diamonds$cut=="Ideal", labels=c("Less than Ideal", "Ideal"))
table(diamonds$ideal, diamonds$cut)
```

# Step 3: Create appropriate graphics
* Create one bivariate graph for each of the four combinations listed in step 1. 
* You can do any combination of the 4 types of bivariate graphs for a total of 4. 
    - Do as many different combinations as you can. Not just 4 graphics on one combination (i.e. C~C)
* For each graph, you need to explain what association (relationship) you are graphing. 
    - Then use the graph to explain the relationship or trends you see in the data. 
    - Place discussion of relationship below the graph it corresponds with.
    - Use specific features of the graphic in your text explanation.
    - Calculate and use summary statistics and also use those in your text explanation. 

# Step 4: Knit to WORD or PDF. Submit to Blackboard by the deadline. 

----

# Example


```{r}
ggplot(diamonds, aes(x=carat, y=price)) + geom_point() + theme_bw() + geom_smooth()
cor(diamonds$carat, diamonds$price)
```

There is a steep positive relationship between the carat of a diamond and it's price. Carat seems to be rounded up a bit, there is very little data in the high decimal places for carats (i.e. between 1.8 and 2.0 carats, between 1.6 and 3.0 carats). This trend sharply levels off around a carat of 2.5, but this dip in the trendline is likely due to the comparativly small number of diamonds over 2.5 carats. Further analysis may restrict the data to diamonds with carats less than 2.5 or 3. Carat and price are also highly correlated, the correlation coefficient is 0.92. This is somewhat unexpected due to the large spread of price data around the trendline for carats over a value of 1. 



