---
title: "Create your own exam question"
output:
  html_document:
    highlight: tango
    theme: cerulean
css: ../css/customh5.css
---

# Instructions
* This is an individual assignment. 
* You are not allowed to use the example I am using in this example file
* Upload the raw markdown file to Blackboard by the due date. 
    - I will change the path to where my raw data is at, and compile your code. If it does not run, you get 0 credit. 

## 1. Import the raw AddHealth data, load any necessary (and only necessary) libraries. 
```{r}
rawdata <- read.csv("E:/MATH315/Project/addHealth_Wave_IV.csv", header=TRUE)
```

## 2. Perform _the minimal amount_ of data management needed for this assignment.
* Explain the variables you are using
```{r}
rawdata$female <- rawdata$BIO_SEX-1
rawdata$female[rawdata$BIO_SEX==6] <- NA
rawdata$BMI <- ifelse(rawdata$H4BMI < 100, rawdata$H4BMI, NA)
```

> For this assignment I am using the following variables
> * `gender` as a binary variable with  0=Male and 1=Female
> * `BMI` as a continuous variable

## 3. Choose any bivariate or multivariate analysis that we have done in the last half of the class. 
* Two sample T-test for a difference in means.
* Two sample T-test for a difference in proportions.
* ANOVA for a difference between many means.
    - Consider following up a significant result with a post-hoc test. 
* Chi-squared test for difference between many proportions.
    - Consider following up a significant result with a post-hoc test. 
* Correlation analysis
* Linear regression analysis
* Logistic regression analysis
* Moderation Analysis
* Confounding Analysis

**1. Define your response and explanatory variable(s)**
**2. State your research question**
**3. State the statistical analysis procedure you will use **

> 1. `BMI` is my continuous response variable, and `gender` is my categorical explanatory variable
> 2. I am testing if the mean BMI is equal between males and females
> 3. I will use a Two-sample t-test for a difference in means.

## 4. Write the code to conduct this analysis. 
```{r}
t.test(BMI~gender, data=rawdata)
```


## 5. Write two multiple choice questions about this analysis. 
* State which answer is correct
* Expain why your answer is right, and your wrong answers are wrong. 
* The questions should address an important concept, not something as trivial as "what's the p-value". 
    - A good exam question tests for **comprehension, or interpretation, not recognition**. 
* The questions should each have a single, clear, correct answer. 
* Eliminate excessive verbiage or irrleevant information from the question. 

> 1. 




