---
title: "Create your own exam question"
output:
  html_document:
    highlight: tango
    theme: cerulean
css: ../css/customh5.css
---

# Instructions
* This is an individual assignment. 
* You are not allowed to use the example I am using in this example file. 
* Upload the **raw markdown file** to Blackboard by the due date. 
    - I will change the path to where my raw data is at, and compile your code. If it does not run, you get 0 credit.
* Scored on a 0-5 scale, based on quality of question and answers. Half credit if you don't explain your answers. 


## 1. Import the raw AddHealth data.
Include any _necessary_ libraries (i.e. `fiefer`)
```{r}
rawdata <- read.csv("J:/MATH315/Project/addHealth_Wave_IV.csv", header=TRUE)
```

## 2. Perform _the minimal amount_ of data management needed for this assignment.
* Explain the variables you are using
```{r}
rawdata$female <- rawdata$BIO_SEX-1
rawdata$female[rawdata$BIO_SEX==6] <- NA
rawdata$BMI <- ifelse(rawdata$H4BMI < 100, rawdata$H4BMI, NA)
```

> For this assignment I am using the following variables
>
> * `female` as a binary variable with  0=Male and 1=Female
> * `BMI` as a continuous variable

## 3. Choose any bivariate or multivariate analysis that we have done in the last half of the class. 
* Two sample T-test for a difference in means.
* Two sample T-test for a difference in proportions.
* ANOVA for a difference between many means.
    - Consider following up a significant result with a post-hoc test. 
* Chi-squared test for difference between many proportions.
    - Consider following up a significant result with a post-hoc test. 
* Correlation analysis
* Linear regression analysis
* Logistic regression analysis
* Moderation Analysis
* Confounding Analysis

**1. Define your response and explanatory variable(s)**  
**2. State your research question**  
**3. State the statistical analysis procedure you will use **  

> 1. `BMI` is my continuous response variable, and `gender` is my categorical explanatory variable  
> 2. I am testing if the mean BMI is equal between males and females  
> 3. I will use a Two-sample t-test for a difference in means.  

## 4. Write the code to conduct this analysis. 
```{r}
t.test(BMI~female, data=rawdata)
```


## 5. Write 1 multiple choice questions about this analysis. 
* Write 1 correct response and 3 incorrect responses (4 choices total)
* State which answer is correct
* Explain why your correct answer is correct, and your wrong answers are wrong. 
* The questions should address an important concept, not something as trivial as "what's the p-value". 
    - A good exam question tests for **comprehension, or interpretation, not recognition**. 
* The questions should each have a single, clear, correct answer. 
* Eliminate excessive verbiage or irrelevant information from the question. 

> **Choose the correct interpretation of this analysis**  
> 
> * A) On average females may have lower BMI compared to males (p=.06).  
> * B) Females have, on average, 0.4 greater BMI compared to males (29.3 vs 28.9, 95% CI for the difference -0.8, 0.2).  
> * C) The average BMI significantly differs between males and females because the confidence interval for the differenc covers 0.   
> * D) I can be 95% confident that the average BMI for males compared to females is (28.9, 29.3).   


Answers

> * A) Incorrect. Females are group 1, which has a higher BMI than males in group 0  
> * B) Correct. The difference in means is 0.4, the direction is correct, and the 95% Interval was reported correctly.   
> * C) Incorrect. If the interval for the difference covers zero, then there is no differnece in means.   
> * D) Incorrect. Those values reported are the group means, not a confidence interval.   



