---
title: "Chapter 5 Notes"
author: "MATH 315"
date: "October 8, 2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(ggplot2)

marriage <- read.csv("J:/MATH315/Data/ageAtMar.txt", sep="")
email <- read.delim("J:/MATH315/Data/email.txt", header=TRUE, stringsAsFactors = FALSE, sep="\t")
```

# Section 5.2: Using R for Inference

### Ex: 5.1 Age at Marriage
According to Wikipedia, the average age for a woman in the US to get married is
28 years (http://en.Wikipedia.org/wiki/Age_at_first_marriage). The average
age at first marriage of 5,534 US women who responded to the National Survey of
Family Growth (NSFG) conducted by the CDC in the 2006 and 2010 cycle was 23.4

**Is there reason to believe that women who respond to the NSFG survey marry
significantly earlier than the average woman?**


I. 


II. 


III. 
```{r, fig.height=4, fig.width=4}
ggplot(marriage, aes(x=age)) +
geom_histogram(aes(y=..density..), color="black", fill=NA) +
stat_function(fun = dnorm, color="red",
args = list(mean = mean(marriage$age), sd = sd(marriage$age)))
qqnorm(marriage$age)
qqline(marriage$age, col="red")
```

IV. 
```{r}
t.test(marriage$age, mu=28, alternative="less")
```



## 5.2.1 Calculating Confidence Intervals using R
```{r}
t.test(marriage$age, mu=28, alternative="two.sided")
```

## 5.3.2 Inference for a proportion using R 

### Example 5.2: Proportion of spam emails - summary data only
Consider the email data set that contains information about the contents in email,
and whether or not the email was flagged as spam. It was found that 367 out of the
3921 emails were flagged as spam. Can we say that the proportion of emails 
fl agged as spam is less than 10%? **What is the sample proportion of spam?**



I. 


II. 


III. 
```{r}
prop.test(x=367, n=3921, p=.1, alternative="less")
```

IV. 


V. 



**Construct a 90% CI for the true proportion of spam. Interpret in context of the problem.**
```{r}
prop.test(x=367, n=3921, p=.1, alternative="two.sided")
```


### Example 5.2: Proportion of spam emails - full data available. 

```{r}
table(email$spam)
```

```{r}
t.test(email$spam, mu=.1, alternative="two.sided", correct=FALSE)
```

### Problem set 5.2: 
```{r}
email$anynum <- ifelse(email$number %in% c("big", "small"), 1, 0)
```




