---
title: 'Cleaning the Depression data set'
date: "Last Updated `r Sys.time()`"
output:
  html_document:
    highlight: tango
    theme: flatly
    toc: yes
    toc_float: yes
---

```{r load2,echo=FALSE, warning=FALSE, message=FALSE}
library(knitr)
#library(rmarkdown)
opts_chunk$set(warning=FALSE, message=FALSE) 
```

# Purpose
Clean the depression data set. 
Codebook can be found [[here]](https://norcalbiostat.netlify.com/data/DepressCodebook.txt)
An annotated version of this data management file can be found [[here]](https://norcalbiostat.github.io/MATH615/docs/lec02_doing_dm.html)


## Import data
```{r, warning=FALSE, message=FALSE}
library(ggplot2)  
depress <- read.table("https://norcalbiostat.netlify.com/data/Depress.txt", 
                      sep="\t", header=TRUE)  
```

## Look at the data

```{r}
str(depress)
```


## Missing data 
Look for missing values
```{r}
table(depress$RELIG, useNA="always")
```

Change 6 to `NA`.
```{r}
depress$RELIG[depress$RELIG==6] <- NA
```


Confirm recode.
```{r}
table(depress$RELIG, useNA="always")
```

```{r}
table(depress$RELIG)
```

Create a continuous measurement
```{r}
y <- c(1, 2, 3, NA, 4, NA, 6)
y
```

try to calculate summary stats with missing data
```{r}
mean(y)
summary(y)
mean(y, na.rm=TRUE)
```



## Detecting and recoding outliers and/or errors

```{r, fig.width=10, fig.height=6}
par(mfrow=c(1,2)) # sets the grahpics grid 1 row by 2 columns
boxplot(depress$AGE)
hist(depress$AGE)
```


```{r}
summary(depress$AGE)
```


```{r}
depress$AGE <- ifelse(depress$AGE==9, 19, depress$AGE)
```

```{r}
summary(depress$AGE)
```




## Identifying variable types (and fixing them)

* What data type does R see this variable as? 
```{r}
table(depress$MARITAL)
str(depress$MARITAL)
is(depress$MARITAL)
```


Turn into a factor
```{r}
depress$MARITAL <- factor(depress$MARITAL, 
                          labels = c("Never Married", "Married", "Divorced", "Separated", "Widowed"))
```

Confirm
```{r}
table(depress$MARITAL)
is(depress$MARITAL)
```

## Creating secondary variables


### Collapsing variables into fewer categories

Collapse into a temp variable
```{r, message=FALSE, warning=FALSE}
library(car)
MARITAL2 <- recode(depress$MARITAL, "'Divorced' = 'Sep/Div'; 'Separated' = 'Sep/Div'")
```

Confirm
```{r}
table(depress$MARITAL, MARITAL2, useNA="always")
```

Apply for real. 
```{r}
depress$MARITAL <- recode(depress$MARITAL, "'Divorced' = 'Sep/Div'; 'Separated' = 'Sep/Div'")
rm(MARITAL2)
```

### Binning a continuous variable into categorical ranges. 

```{r}
depress$inc_cut <- Hmisc::cut2(depress$INCOME, cuts=c(30,40,50,60))
table(depress$inc_cut)
```

### Dichotomizing
Dichotomous variables tend to be binary indicator variables where a code of `1` is the level you're interested in. 

```{r}
depress$SEX <- depress$SEX -1 
table(depress$SEX)
```


```{r}
mean(depress$SEX)
```

```{r}
table(depress$DRINK)
```

```{r}
depress$DRINK <- ifelse(depress$DRINK==2, 0, depress$DRINK)
table(depress$DRINK)
``` 




###  Sum or Average values across multiple variables

```{r}
depress$sleep <- depress$C5 + depress$C11 + depress$C19
```

```{r}
summary(depress$sleep)
```

# Renaming varible names for sanity sake

```{r}
names(depress) <- tolower(names(depress))
```


# Saving your changes

```{r, eval=FALSE}
date <- format(Sys.Date(), "%m%d%y")
filename <- paste("depress_", date, ".txt", sep="")
write.table(depress, filename, sep="\t", row.names=FALSE)
```

